The variability of Betelgeuse explained by surface convection

### 2024-05-06 main.tex
Intro proofed

\section{Introduction}
Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods can be found in the 
literature usually clustered around the so-called long secondary period (LSP) of 2000 days, and other shorter periods of 400 and 200 days \citep{kiss_variability_2006}. The origin of the LSP remains poorly known, some tried to explain the LSP by the lifetime of the large convective cells present at the surface of RSG (e.g \cite{stothers_giant_2010}) or by the magnetic field \citep{wood_long_2004}. The shorter periods of 400 and 200 days seem to have a different origin than the LSP. By studying a large sample of RSG, \cite{kiss_variability_2006} attributed the 400 days period to radial pulsation modes, the 200 days period being attributed to the first overtone \citep{joyce_standing_2020}, which was in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

Besides radial pulsation, surface activity leading to random brightness variation was also mentioned to explain the variability of Betelgeuse, being firstly theorised by \cite{schwarzschild_scale_1975} and \cite{gray_mass_2008} proposed that the 400 days period was a consequence of the convective activity. Betelgeuse is known to present large convective cells with lifetimes in the order of one to two years \citep{lopez_ariste_convective_2018} and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns \citep{gray_mass_2008}. \\

@@

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the great dimming \citep{guinan_fall_2020}. 
From interferometric data, it is speculated that this event was caused by a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, that might be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other theories such as a drop in temperature was also mentioned by \cite{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event it turns out that a change in brightness was not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continue its random variation of brightness. 
Interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed after the dimming. Using the light curves from AAVSO, 
the authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period, while after the great dimming the main periods of Betelgeuse have shortened, oscillating between 97 days and 230 days \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, it brings up the question whether all changes in brightness cannot be due to similar changes in the brightness 
distribution over the disk, unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question we shall seek for such typical periods of 400 and 200 days in  
observational proxies related to the convective activity.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

Linear polarization in the atomic lines of the spectra of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus it must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different height of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with contemporaneous images made with interferometric 
techniques \citep{montarges_close_2016} and show clear convective patterns, akin to solar granulation.\

Linear polarization observed in the atomic lines is therefore a proxy of convection, unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the aforementioned periods 
in linear polarization, we must conclude that these periods are related to the convective activity which originates those linear polarization signals.
%and unrelated to any radial pulsation phenomena. 
This is the purpose of the present work. In section \ref{Section 2} we describe the dataset of linear polarization obtained with the TBL and the Lomb-Scargle periodogram used to derive periods. In section \ref{section 3} we associate the 200 days period to convective timescale of the smallest granules and the 330 days periodicity to the largest granules, and we provide an explanation in the change of variability of Betelgeuse before and after the great dimming. In section \ref{Conclusion}, we summup our work.

>>>>

\section{Introduction}
Betelgeuse is a prototypical red supergiant (RSG) known to be a semi-regular variable. Several periods for this variability can be found in the 
literature, usually clustered around the so-called long secondary period (LSP) of 2000 days, plus shorter periods of 400 and 200 days \citep{kiss_variability_2006}. 
The origin of the LSP remains poorly known: some have tried to find the explanation of  the LSP in the lifetime of the large convective cells present at the surface of RSG (e.g \cite{stothers_giant_2010}) while others have invoked a the magnetic field \citep{wood_long_2004}. 
The shorter periods of 400 and 200 days seem to have a different origin than the LSP. By studying a large sample of 
RSG, \cite{kiss_variability_2006} attributed the 400 days period to radial pulsation modes, while the 200 days period was attributed to the first overtone \citep{joyce_standing_2020}, 
 in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

Besides radial pulsation, surface activity leading to random brightness variation has also been mentioned to explain the variability of Betelgeuse. \textbf{Having been firstly theorised by \cite{schwarzschild_scale_1975} and \cite{gray_mass_2008} 
proposed that the 400 days period was a consequence of the convective activity}. Betelgeuse is known to present large convective cells with lifetimes in the order of one to two years \citep{lopez_ariste_convective_2018} and measurable changes 
in the span of one week. Bright convection cells near disk center will increase the integrated brightness of the star when 
compared with other situations where such cells are found near the edges. So even without calling for the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create a variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns \citep{gray_mass_2008}. \\

@@

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the Great Dimming \citep{guinan_fall_2020}. 
From interferometric and spectroscopic data, it has been determined that this event was caused by a cloud of dust formed close to the line of 
sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, that might be caused by a 
mass loss event and leading to this dimming \citep{dupree_great_2022}. Other hypothesis have  also been explored , such as a drop in temperature \citep{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. 
This event is an extreme illustration of  changes in brightness that were not due to any kind of pulsation but to a change on the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continued its random variation of brightness. 
But quite interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed since the dimming. Using the light curves from AAVSO, 
the cited authors showed that before the dimming, the dominant period of Betelgeuse was the 400 days period, while after the great dimming the main periods of Betelgeuse have shortened, oscillating between 97 days and 230 days \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, such modifications of the variability periods  bring up the question whether all changes in brightness cannot be due to similar changes 
in the brightness  distribution over the disk, and unrelated to any pulsation phenomenon. 
%Our interest in an alternative explanation of the variability
%in terms  of convective patterns arose. 
In order to address this question we  seek for the typical periods of 400 and 200 days in  
observational proxies which are related to the convective activity but not to any pulsation, such as the linear polarization spectra.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

Linear polarization in the atomic lines of the spectrum of Betelgeuse, discovered by \cite{auriere_discovery_2016}, has been interpreted as the joint action of 
two mechanisms. First the depolarization of the continuum by atoms which absorb linearly polarized light from the continuum and re-emit unpolarized light, the 
continuum photons being polarized by Rayleigh scattering. This depolarization produces signals with an azimuth symmetry over the disk that 
would  cancel out the net linear polarization on a homogeneous disk. Thus, such mechanism of production of polarization must be combined with an inhomogenous disk to produce the net linear polarization 
signal observed. This interpretation suggested the possibility of mapping those brightness inhomogeneities. This has been done by \cite{lopez_ariste_convective_2018} 
even producing 3-dimensional images of the atmosphere of Betelgeuse \citep{lopez_ariste_three-dimensional_2022} when taking advantage of the different heights of formation of 
different lines in the spectrum of Betelgeuse. The images produced compare very well with contemporaneous images made with interferometric 
techniques \citep{montarges_close_2016} and show clear convective patterns, akin to solar granulation.

Linear polarization observed in the atomic lines is therefore a proxy of convection, unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the 
aforementioned variability periods 
in linear polarization, we must conclude that these periods are related to the convective activity which originates the linear 
polarization signals.
%and unrelated to any radial pulsation phenomena. 
This is the purpose of the present work.

In section \ref{Section 2} we describe the dataset of linear polarization obtained with the Telescope Bernard Lyot (TBL) and the Lomb-Scargle technique used to derive periods. 
In section \ref{section 3} we associate the 200 days period to convective timescale of the smallest granules and the 330 days periodicity to the largest granules, and we provide an explanation in the change of variability of Betelgeuse before and after the great dimming. 
In section \ref{Conclusion}, we summup our work.
